MinCall --- MinION end2end deep learning basecaller

### 2017-07-01 paper/main.tex
Passed through grammar check

    The best performing deep learning model achieves state-of-the-art median 91.4\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
\keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
\end{abstract}

@@

Although MinION is able to produce long reads, even up to 882 kb~\cite{loman1-100k,loman2-800k}, they have a high sequencing error rate. This has been somewhat alleviated with new R9 pore model, replacing older R7 ones. In this paper, we show that this error rate can be reduced by our approach with properly trained neural network model.

The exact error rate metric is unreliable since multiple pipeline tools could be the issue. First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data. Next, our model (or any other) comes along and basecalls the sequences. To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, most commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.

@@

We compared our model on R9 chemistry with Metrichor (HMM based approach), Nanonet and DeepNano (RNN based approaches). For detailed basecaller overview see the appendix~\ref{app:basecallers}.

@@

Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc. dr. sc. Petra Korać i dr.sc. Paula Dobrinić}. Both used datasets show in table \ref{tbl:datasets} have been previously passed through MinKNOW and had been basecalled by Metrichor. As 1D read analysis was the focus of this paper, only those reads were used.

@@

\section{Method}
Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}. For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence. Other building blocks used are Batch normalization(BN)~\cite{BNORM} and pooling layers. Dropout~\cite{srivastava2014dropout} was not used.
Described model is implemented using tensorflow~\cite{tensorflow2015-whitepaper} and source warp-ctc~\cite{warpctc} GPU CTC loss implementation.

@@

During testing, we tried ReLu and PrELU~\cite{prelu} with no significant result difference. We also tried different channel numbers, receptive field width, and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly. We used SigOpt~\cite{dewancker2016strategy}\footnote{\url{https://sigopt.com/}} bayesian hyperoptimization library.

@@

In Appendix~\ref{app:evaluation:CIGAR} the histogram figure~\ref{fg:cigar_op_dist} shows how matches, mismatches, insertions, and deletions are distributed across the reads. It is shown that mismatches and insertion occur more frequently at the beginnings and the ends of the reads. This is not only the case for the developed basecaller, but all other show the same property. This could be due to lack of context information from both sides when edges are basecalled.

@@

All models show a slight bias towards deletions than insertions, but this may be the limitation of technology as it has been reported that deletion and mismatch rates for nanopore data are ordinarily higher than insertion rates~\cite{sovic2016fast}. Results are shown in table~\ref{tbl:spec_ecoli}. Lambda table is deferred to appendix in table~\ref{tbl:spec_lambda}.

@@

During this paper creation we used Sigopt academic licence for hyperparameter optimization. We gratefully acknowledge the support of NVIDIA Corporation with the donation of the Titan Xp GPU used for this research.

@@

For data, we're thankful for cooperation of Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić and Loman labs~\cite{loman1-100k} for publicly posting their dataset.

@@

With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation. Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet and basecaller integrated into MinKNOW \cite{ont-basecallers}.

\emph{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
It is available to the Nanopore Community served as a binary. The source code of Albacore was not provided and is only available through the ONT Developer Channel. Tool supports only R9.4 and future R9.5 version of the chemistry. For R9 tests in this paper we used R9.4 chemisty setting as instructed on ONT forums.

\emph{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use. It uses \emph{CURRENNT} library for running neural networks. It supportes basecalling of both R9 and R9.4 chemistry versions. However in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.

\emph{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies. Similar to Nanonet, it is the platform for ongoing development. Scrappie is reported to be the first basecaller  that specifically address homopolymer basecalling. It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.

>>>>

    The best performing deep learning model achieves state-of-the-art  91.4\% median match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
\keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
\end{abstract}

@@

Although MinION is able to produce long reads, even up to 882 kb~\cite{loman1-100k,loman2-800k}, they have a high sequencing error rate. This has been somewhat alleviated with new R9 pore model, replacing older R7 ones. In this paper, we show that this error rate can be reduced by our approach with the properly trained neural network model.

The exact error rate metric is unreliable since multiple pipeline tools could be the issue. First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible then sequenced using the MinION device obtaining raw data. Next, our model (or any other) comes along and basecalls the sequences. To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, most commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.

@@

We compared our model on R9 chemistry with Metrichor (HMM-based approach), Nanonet and DeepNano (RNN based approaches). For the detailed basecaller overview see the appendix~\ref{app:basecallers}.

@@

Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc. dr. sc. Petra Korać and dr.sc. Paula Dobrinić}. Both used datasets show in table \ref{tbl:datasets} have been previously passed through MinKNOW and had been basecalled by Metrichor. As 1D read analysis was the focus of this paper, only those reads were used.

@@

\section{Method}
Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}. For loss, we used CTC (Connectionist Temporal Classification)~\cite{graves2006connectionist} between basecalled and the target sequence. Other building blocks used are Batch normalization(BN)~\cite{BNORM} and pooling layers. Dropout~\cite{srivastava2014dropout} was not used.
Described model is implemented using tensorflow~\cite{tensorflow2015-whitepaper} and source warp-ctc~\cite{warpctc} GPU CTC loss implementation.

@@

During testing, we tried ReLu and PrELU~\cite{prelu} with no significant result difference. We also tried different channel numbers, receptive field width, and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly. We used SigOpt~\cite{dewancker2016strategy}\footnote{\url{https://sigopt.com/}} bayesian hyperparameter optimization library.

@@

In Appendix~\ref{app:evaluation:CIGAR} the histogram figure~\ref{fg:cigar_op_dist} shows how matches, mismatches, insertions, and deletions are distributed across the reads. It is shown that mismatches and insertion occur more frequently at the beginnings and the ends of the reads. This is not only the case for the developed basecaller, but all other shows the same property. This could be due to lack of context information from both sides when edges are basecalled.

@@

All models show a slight bias towards deletions than insertions, but this may be the limitation of technology as it has been reported that deletion and mismatch rates for nanopore data are ordinarily higher than insertion rates~\cite{sovic2016fast}. Results are shown in table~\ref{tbl:spec_ecoli}. Lambda table is deferred to Appendix in table~\ref{tbl:spec_lambda}.

@@

During this paper creation, we used Sigopt academic license for hyperparameter optimization. We gratefully acknowledge the support of NVIDIA Corporation with the donation of the Titan Xp GPU used for this research.

@@

For data, we're thankful for the cooperation of Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić and Loman labs~\cite{loman1-100k} for publicly posting their dataset.

@@

With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation. Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet, and basecaller integrated into MinKNOW \cite{ont-basecallers}.

\emph{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
It is available to the Nanopore Community served as a binary. The source code of Albacore was not provided and is only available through the ONT Developer Channel. The tool supports only R9.4 and future R9.5 version of the chemistry. For R9 tests in this paper, we used R9.4 chemistry setting as instructed on ONT forums.

\emph{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use. It uses the \emph{CURRENNT} library for running neural networks. It supports basecalling of both R9 and R9.4 chemistry versions. However, in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.

\emph{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies. Similar to Nanonet, it is the platform for ongoing development. Scrappie is reported to be the first basecaller  that specifically address homopolymer basecalling. It became publicly available just recently in June 2017 and supports R9.4 and future R9.5 data.
